\chapter{Review \#1}
\label{ch:rev1n}
\begin{fullwidth}

\section{List of Topics}
\begin{enumerate}
\item Mathematical preliminaries:
\begin{enumerate}
\item Number representation.
\begin{enumerate}
\item Unsigned integer representations.
\item IEEE-754 floating point representation.
\end{enumerate}
\item Sources of error.

\item Background of Linear Algebra.
\begin{enumerate}
\item matrix/vector operations.
\item matrix properties.
\end{enumerate}
\end{enumerate}

\item Solution of Non-linear Equations.
\begin{enumerate}
\item Bisection method.
\item Newton's method (single equation and system of equations).
\item Secant method.
\item Steffen's method.
\item MATLAB built-in methods (single equation and system of equations).
\end{enumerate}

\item Solution of Linear System of Equations.

\begin{enumerate}
\item Gauss elimination with and without pivoting.
\item LU factorization with and without pivoting.
\item MATLAB built-in methods.
\item Error bounds for solution of linear systems of equations.
\item Iterative methods:
\begin{enumerate}
\item Jacobi, Gauss-Seidel, SOR.
\item MATLAB built-in methods including: \lstinline[style=myMatlab]{pcg} and \lstinline[style=myMatlab]{gmres} with preconditioning.
\end{enumerate}
\end{enumerate}

\end{enumerate}

\section{Review Questions}

\begin{enumerate}
\item Write down the sign bit, exponent bits, and first 6 bits of the mantissa for IEEE-754 double precision floating point representation of the number 2023.

\vspace{5.0cm}

\item Write down the number 81 in 32-bit unsigned integer format (little-endian layout).

\vspace{5.0cm}

\item Carry out 2 iterations of the secant method to find the root of $f(x) = x - 2e^{-x}$ starting with $x_1 = 0$ and $x_2 = 1$. (i.e. find $x_3$ and $x_4$)

\vspace{15.0cm}

\item Consider the following system of nonlinear equations:
\begin{align*}
-2x^3 + 3y^2 + 42 &= 0 \\
5x^2 + 3y^3 - 69 &= 0
\end{align*}
Using Newton's method for a system of equations and starting at $x=y=1$ write the linear system of equations you would need to solve to find $\Delta x$ and $\Delta y$ to update the estimated solution.


\vspace{4.0cm}

\item Using MATLAB's built-in function: \lstinline[style=myMatlab]{[L,U,P] = lu(A)} function, find the solution to the linear system of equations shown below:

\begin{equation*}
\bracketMatrixstack{
0 & 3 & 8 & -5 & -1 & 6 \\
3 & 12 & -4 & 8 & 5 & -2 \\
8 & 0 & 0 & 10 & -3 & 7 \\
3 & 1 & 0 & 0 & 0 & 4 \\
0 & 0 & 4 & -6 & 0 & 2 \\
3 & 0 & 5 & 0 & 0 & -6 
}
\bracketVectorstack{
x_1 \\
x_2 \\
x_3 \\
x_4 \\
x_5 \\
x_6
}
=
\bracketVectorstack{
34 \\
20 \\
45 \\
36 \\
60 \\
28
}
\end{equation*}
\end{enumerate}

\end{fullwidth}
