\chapter{Assignment \#6}
\label{ch:ass6}
\begin{fullwidth}

Show that the given functions are orthogonal on the given interval.
\begin{enumerate}
\item $f_1(x) = e^{x}, \ f_2(x)=xe^{-x}-e^{-x}, \ \ x\in[0,2]$

\vspace{1.0cm}

\item $f_1(x) = x, \ f_2(x)=\cos{2x}, \ \ x\in\left[-\sfrac{\pi}{2},\sfrac{\pi}{2}\right]$
\end{enumerate}

\vspace{1.50cm}

\noindent Show that the given set of functions is orthogonal on the indicated interval.  Find the norm of each function in the set.
\begin{enumerate}[resume]
\item $\left\{\sin{x},\sin{3x},\sin{5x},\dots \right\}, \ \ x\in\left[0,\sfrac{\pi}{2}\right]$

\end{enumerate}

\vspace{1.5cm}

\noindent Use MATLAB to verify by numeric integration that the functions are orthogonal with respect to the indicated weight function on the given interval:

\begin{enumerate}[resume]
\item $H_0(x) = 1, \ H_1(x) = 2x, \ H_2(x)=4x^2-2; \ \ w(x)=e^{-x^2}, \ \ x\in(-\infty,\infty)$

\noindent\textbf{Note:} use the built-in function \lstinline{integral()} to do the numeric integration.  In MATLAB, $-\infty$ and $\infty$ are represented by \lstinline{-inf} and \lstinline{inf} respectively.
\end{enumerate}


\vspace{1.5cm}

\noindent Use MATLAB to construct the Fourier series expansion of the given function $f(x)$ on the given interval.  For each problem create a plot that shows: a) $f(x)$ along with b) the truncated Fourier series of $f(x)$ with \lstinline{N=5} and c) \lstinline{N=15} terms.  Also give the number to which the Fourier series expansion converges at any point(s) of discontinuity in $f(x)$. 

\begin{enumerate}[resume]
\item $f(x) = 
\begin{cases}
0, & -\pi < x < 0 \\
1, & 0 \le x < \pi
\end{cases}$


\vspace{1.0cm}

\item $f(x) = 
\begin{cases}
0, & -\pi < x < 0 \\
x^2, & 0 \le x < \pi
\end{cases}$

\vspace{1.0cm}

\item $f(x) = 
\begin{cases}
0, & -2 < x < 0 \\
-2, & -1 \le x < 0 \\
1, & 0 \le x < 1 \\
0, & 1 \le x < 2
\end{cases}$

\end{enumerate}

\vspace{1.5cm}

\noindent Determine whether the given function is even, odd, or neither.
\begin{enumerate}[resume]
\item $f(x) = x^2+x$

\vspace{1.0cm}

\item $f(x) =
\begin{cases}
x^2, & -1 < x < 0 \\
-x^2, & 0 \le x < 1
\end{cases}$
\end{enumerate}

\vspace{1.5cm}

\noindent Use MATLAB to expand the given function in an appropriate cosine or sine series.  For each function create a plot showing: a) $f(x)$ along with b) the truncated Fourier series of $f(x)$ with \lstinline{N=5} and; c) \lstinline{N=15} terms.

\begin{enumerate}[resume]
\item $f(x) = \left| x \right|, \ \ -\pi < x \pi$

\vspace{1.0cm}

\item $f(x) = 
\begin{cases}
x-1, & -\pi < x < 0 \\
x+1, & 0 \le x < \pi
\end{cases}$

\vspace{1.0cm}

\item $f(x) = 
\begin{cases}
1, & -2 < x < -1 \\
-x, & -1 \le x < 0 \\
x, & 0 \le x < 1 \\
1, & 1 \le x < 2
\end{cases}$

\vspace{1.0cm}

\item $f(x) = 
\begin{cases}
x, & 0 < x < \sfrac{\pi}{2} \\
\pi - x, & \sfrac{\pi}{2} \le x < \pi
\end{cases}$
\end{enumerate}

\end{fullwidth}
